---
title: "Data frames (basics)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup,include=FALSE,eval=TRUE}
source( "../R_course_setup.R" );
```

# Data frames (basics)

So we are now able to read in data sets (tables). A data set in R is called a `data.frame`. 

```{r result="asis", echo=FALSE, comment="", eval=TRUE}
pulse <- catReadTable( "pulse.txt", "pulse" )
```
```{r eval=FALSE}
pulse <- read.delim( "../data/pulse.txt" )
```

```{r result="asis", echo=FALSE, comment="", eval=TRUE}
survey <- catReadTable( "survey.txt", "survey" )
```
```{r eval=FALSE}
survey <- read.delim( "../data/survey.txt" )
```

R can have many data sets in memory simultaneously. You will always have to specify which data set you are working in.

## Exploring 

### Dimensions

Rows in a data.frame are typically subjects; columns are variables.

To find the size of a data.frame
```{r}
ncol(pulse)
nrow(pulse)
dim(pulse)
```

### Showing head and tail

To get a quick impression of a data.frame:

- `head` prints the first 6 rows
- `tail` prints the last 6 rows

If you want more or less than 6, add the number you want as a second argument to the function:
```{r}
head(pulse)
tail(pulse, 3)
```

We've already seen `View`.

With `names` (alternatively `colnames`) you find the names of the variables (columns) in the data.frame:
```{r}
names(pulse)
```

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "explore", "04_dataframes.Rmd" );
```


## Extracting 

### Columns

To extract a column of a data.frame use `$`. The result is a vector:
```{r}
pulse$age
```

Note the autocomplete in RStudio that helps suggest the right column.

To add a column to a data.frame, use the assigmnent:
```{r}
pulse$pulse.diff <- pulse$pulse2 - pulse$pulse1
head( pulse )
```


### Row names

A data.frame always has row names. Note that these names are not themselves a column of the data.frame!
```{r}
rownames(pulse)
```

Name of each row must be unique.

### Elements

An individual entry to a data.frame can be extracted using square brackets `[`, either using the names of row and column (note the quotes) or their indices. Row comes before the comma, column after.
```{r}
pulse["1993_E", "height"]
pulse[5, 2]
```

You can also use ranges like with vectors
```{r}
pulse[4:6, 1:5]
```

Much more about using square brackets later in the course.


### (*) Removing a column 

To remove a column from a data.frame, assign `NULL` to that column:
```{r}
pulse$pulse.diff <- NULL
```

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "columns", "04_dataframes.Rmd" );
```

## Example data

R contains many example data sets. To see which, see:
```{r, eval=FALSE}
data()
```

Example data are immediately accessible in R. For example:
```{r}
BOD
```

Some description is always available:
```{r, eval=FALSE}
?BOD
help("BOD")
```

We will make use of example data from the `MASS` package. Packages are bundles with additional functions and data. To make `MASS` available in your R session, say:
```{r}
library(MASS)
```
More about packages later.

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "newFrame", "04_dataframes.Rmd" );
```
