---
title: "Basic data types"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup,include=FALSE,eval=TRUE}
source( "../R_course_setup.R" );
```

# Basic data types

We will use the `pulse` and `survey` data again for illustration and exercises:

```{r result="asis", echo=FALSE, comment="", eval=TRUE}
pulse <- catReadTable( "pulse.txt", "pulse" )
```
```{r eval=FALSE}
pulse <- read.delim( "data/pulse.txt" )
```

```{r result="asis", echo=FALSE, comment="", eval=TRUE}
survey <- catReadTable( "survey.txt", "survey" )
```
```{r eval=FALSE}
survey <- read.delim( "data/survey.txt" )
```

## Types

Columns in a data.frame can be of different types. Typically:

- numeric (we've seen already)
- character (for text)
- factor (for categorical variables)

## `str` and `class`

To get a quick overview (structure) of the types of data in your data.frame:
```{r}
str(pulse)
```

To learn about the type of a specific column:
```{r}
class(pulse$name)
```

# Vector classes

## Numeric data

Numeric data can be `integer` (whole numbers) or `numeric` (continuous data) but you can ignore that distinction if you are not a programmer.

We've seen how to make numeric data with `c` or `:`

### Useful functions for numeric data

Summarizing a single variable:

- `mean`
- `median`
- `min`
- `max`
- `range` (two values: min, max)
- `sd` (standard deviation)
- `var` (variance)
- `hist` (histogram)

A six-number summary: range, three quartiles and the mean:
```{r}
summary(pulse$age)
```

### Relationships between two variables

- `cor` (correlation)
- `plot` (scatterplot)

```{r}
cor(pulse$height, pulse$weight)
plot(pulse$height, pulse$weight)
```


```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "numeric", "05_character_factor_basics.Rmd" );
```

## Character data

Texts in R is called `character`. You recognize is by the quotes around the values.

### Creating character data

Use either single or double quotes

```{r}
text <- c('alpha', 'beta', 'gamma')
TEXT <- c("ALPHA", "BETA", "GAMMA")
```

### Names

Row and column names in a data.frame are always `character`:
```{r}
rownames(pulse)[1:10]
```

### (*) Useful premade character vectors

- `LETTERS` (capitals)
- `letters` (lower case)
- `month.name` (months)

### (*) Calculating with character vectors

Arithmetics of text

```{r}
toupper("me")
paste(LETTERS, letters, sep='_')
```


## Factor - categorical data

A categorical variable in R is called `factor`:

- They are internally coded as numbers 1,2,3,...
- The numbers have value labels attached to them (called levels)

### Making a factor

To make a factor variable, start with a character vector and use `factor`:

```{r}
fac <- factor(c('A', 'B', 'A', 'B', 'B'))
```

### Factor basics

You recognize a factor by the `Levels:` line when printing:
```{r}
fac
```

To get the numeric coding:
```{r}
as.numeric(fac)
```

To get the value labels:
```{r}
levels(fac)
nlevels(fac)
```

### Turning factor (back) into character

```{r}
as.character(fac)
```

### Table

Factors are best summarized with `table`:
```{r}
table(pulse$exercise)
```

`table` can also crosstabulate two (or even more) variables:
```{r}
table(pulse$gender, pulse$exercise)
```

### (*) Table of table

It is suprisingly useful to use table twice.

```{r}
table(table(pulse$name))
```

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "factors", "05_character_factor_basics.Rmd" );
```








