# ------------------------------------------------------------------------------
# Header and title line
# ------------------------------------------------------------------------------

---
title: "The title"
author:
- affiliation: "Biomedical Data Sciences, LUMC"
  company: "LUMC"
  name: "Jelle Goeman"
- affiliation: "Biomedical Data Sciences, LUMC"
  company: "LUMC"
  name: "Ramin Monajemi"
- affiliation: "Biomedical Data Sciences, LUMC"
  company: "LUMC"
  name: "Szymon M. Kie&#322;basa"
date: "`r format(Sys.time(), '%d %B %Y')`"
always_allow_html: yes
---

```{r setup,include=FALSE,eval=TRUE}
source( "../R_course_setup.R" );
```

# The title (again)

# ------------------------------------------------------------------------------
# Using git diff for cooperating Windows, Mac and Linux users
# ------------------------------------------------------------------------------

# git diff --ignore-space-at-eol

# ------------------------------------------------------------------------------
# Loading pulse and survey
# ------------------------------------------------------------------------------

```{r setup,include=FALSE,eval=TRUE}
source( "../R_course_setup.R" );
```

```{r result="asis", echo=FALSE, comment="", eval=TRUE}
pulse <- catReadTable( "pulse.txt", "pulse" )
```
```{r eval=FALSE}
pulse <- read.delim( "data/pulse.txt" )
```

```{r result="asis", echo=FALSE, comment="", eval=TRUE}
survey <- catReadTable( "survey.txt", "survey" )
```
```{r eval=FALSE}
survey <- read.delim( "data/survey.txt" )
```

# ------------------------------------------------------------------------------
