---
title: "Regression and formula objects"
date: "`r format(Sys.time(), '%d %B %Y')`"
---


```{r setup, include=FALSE}
source( "../R_course_setup.R" );
```
```{r echo=FALSE}
pulse <- read.delim("../data/pulse.txt")
survey <- read.delim("../data/survey.txt")
```


# Formula objects

Formula objects are the way to tell R that one variable depends on another. 

## Basics of formula objects

To specify a (statistical) model in which `y` depends on `x`, say

```{r}
y ~ x
```

We use a formula for more readable specification of code. For example, when plotting. Instead of

```{r}
plot(pulse$pulse1, pulse$pulse2)
```

we say

```{r}
plot(pulse2 ~ pulse1, data=pulse)
```

Note the reverse order!

If a function allows a formula as input it always has a `data` argument. This gives the data.frame (or other environment) in which the variables in the formula are interpreted.


## Use of formula objects in statistics

Formula objects can also be used in the syntax of tests and other functions, making them simpler. For example, consider the problem of comparing values of `pulse1` between males and females. We can use an unpaired t test for this, by writing:


```{r}
pulse1.male <- pulse$pulse1[survey$gender == 'male']
pulse1.female <- pulse$pulse1[survey$gender == 'female']
t.test(pulse1.male, pulse1.female)
```

A much simpler way is to use a formula:

```{r}
t.test(pulse1 ~ gender, data = pulse)
```

Many functions allow (or require!) formula as input.


## (*) The formula class

A formula is just a R object.

```{r}
class(y~x)
```

It can be stored in a variable and reused.

```{r}
form <- pulse1 ~ gender
t.test(form, data=pulse)
```


```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "formula", "03_regression.Rmd" );
```



# Simple linear regression

## The `lm` function

The command for linear regression is `lm` (for *linear model*). The linear model returns an object of class `lm`.

```{r}
fit <- lm(weight ~ height, data = pulse)
```

The output of `lm` is an object of class `lm`. 

```{r}
fit
names(fit)
```

Some S3 objects have special functions defined for them. The following functions extract useful information from the `lm` object.

```{r}
summary(fit)
coef(fit)
residuals(fit)
fitted.values(fit)
```

Note that `summary(fit)` returns itself an object in which some additional things are calculated.

```{r}
summa <- summary(fit)
class(summa)
names(summa)
```

Most useful is the regression table

```{r}
coef(summa)
```

and the confidence intervals for all or some regression coefficients

```{r}
confint(fit)
confint(fit, "height")
```


##  Visualizing a regression 

We can easily visualize the regression using the same formula and fit object

```{r}
plot(weight ~ height, data=pulse)
abline(coef(fit))
```


```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "lm", "03_regression.Rmd" );
```



# Multiple regression and prediction

## Multiple terms in a formula

We can have multiple terms in a formula. This way we can do multiple regression

```{r}
fit <- lm(pulse2 ~ pulse1 + height, data=pulse)
fit
summary(fit)
```


## Predicting

We can use regression for prediction using the predict function. To predict we need two things. First, a fitted model object.

```{r}
fit <- lm(pulse2 ~ pulse1 + height, data=pulse)
```

Second, a data.frame with new data for our covariates

```{r}
new.data <- data.frame(pulse1=c(90, 80), height=c(c(173, 185)))
```

Now we can predict a value for a person with these covariates

```{r}
predict(fit, new.data)
```


## (*) Interaction

Specifying interactions between variables in a formula: use `:` or `*`:

- `:` interaction only
- `*` interaction and main effects

Let us say that we want to explain `pulse2` by `pulse1`, `exercise` and `alcohol`, as well as a variable representing an interaction between `exercise` and `alcohol`. Two alternative ways of specifying the same model are

```{r}
lm(pulse2 ~ pulse1 + exercise + alcohol + exercise:alcohol, data=pulse)
lm(pulse2 ~ pulse1 + exercise*alcohol, data=pulse)
```


## (*) The intercept term

We see that R automatically adds an intercept term to the model. You can suppress the intercept too, by adding either `+0` or `-1` to the formula. Suppressing the intercept has different effects if there are factor variables in your model or not. 

Suppression of the intercept means regression through the origin

```{r}
lm(weight ~ 0 + height, data=pulse)
```

Note that this is different if we have factors (see next part)!

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "multiple", "03_regression.Rmd" );
```



