---
title: "Character strings & Pattern matching (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
#pulse <- read.table("https://barmsijs.lumc.nl/R_course/data/pulse.txt", header = T) 
#survey <- read.delim("https://barmsijs.lumc.nl/R_course/data/survey.txt") 
pulse <- read.table("../data/pulse.txt")
survey <- read.delim("../data/survey.txt")
source( "../R_course_setup.R" )
```

# Character strings & Pattern matching (tasks)

## Tasks

### concatenation {#characterVectors}

The `survey` data is used throughout the exercises.

```{r, echo=TRUE, eval=TRUE, comment=NA}
#survey <- read.delim("https://barmsijs.lumc.nl/R_course/data/survey.txt") 
survey <- read.delim("../data/survey.txt")
# Use subject names (column name) in survey data set
surveyNames <- as.vector(survey$name)
# Domain names
domains    <- c("lumc.nl", "leidenuniv.nl", "vumc.nl")
```


1. Generate a vector of e-mail addresses `@leidenuniv.nl` for all names in `surveyNames`.  

```{r}
emails <- paste(surveyNames, "leidenuniv.nl", sep="@")
head(emails)
```

2. Mailing list: create a single string with all e-mail addresses separated by ';'. 

```{r}
mailingList <- paste(emails, collapse = ";")
```

3. Create a mailing list of all names randomly assigned to `domains` (hint: let recycling do the semi-random selection.) 

```{r}
# The vector 'surveyNames' is the largest vector therefore '@' and 'domains' are recycled.  
#
emails      <- paste(surveyNames, domains, sep="@")
mailingList <- paste(emails,collapse = ";")
```


4. Predict the output for the following statements, note that paste0(...) <=> paste(...,sep="") and rev(x) reverses `x`: 

```{r, echo=TRUE}
paste(c("a","b"),"c",letters[1:3], sep="")
paste0(paste0(letters[19:20],collapse = ""),"a", paste0(rev(letters[19:20]), collapse = ""))
```

5. Plot a histogram of name lengths. (hint: ?hist, ?nchar)

```{r}
hist(nchar(surveyNames))
```

### strsplit {#strsplit}

1. Recreate the vector of e-mail addresses  from the *mailingList* ( previous exercise (paste: nr.3) ) using `strsplit(...)` function.

```{r}
emails <- unlist(strsplit(mailingList,";")) 
head(emails)
length(emails)
```

2. Given the vector of e-mail addresses just obtained retrieve the user names only (hint: ?do.call.)

```{r}
usersAndDomains <- do.call(rbind, strsplit(emails, "@"))
colnames(usersAndDomains) <- c("user","domain")
head(usersAndDomains)
userNames <- usersAndDomains[,"user"]
length(userNames)
head(userNames)
```

3. Create a frequency table of the number of users per domain (hint ?`table`).

```{r}
usersAndDomains <- do.call(rbind, strsplit(emails, "@"))
colnames(usersAndDomains) <- c("user","domain")
table(usersAndDomains[,"domain"])
```

### grep/grepl/sub {#grep}

1. Create a list with elements 'lumc', 'leidenuniv' and 'vumc' containing  the corresponding e-mails using the vector `emails` from concatenation (1.1.1) exercise 3. 

```{r}
list(lumc=grep("lumc", emails, value = T), 
     leidenuniv=grep("leidenuniv", emails, value = T),
     vumc=grep("vumc", emails, value = T))
```

2. Create a mailing-list (semi-colon separated) with all user names containing pattern 'in'. 

```{r}
paste(grep("in", emails , value = T), collapse = ";")
```


3. Find the most popular letter which names start with.

```{r}
firstChar <- substr(surveyNames,1,1)
firstCharCounts <- table(firstChar)
( mostPopularFirstChar <-  names(firstCharCounts[firstCharCounts==max(firstCharCounts)]) )
```

4. Find the names that start with the most popular first letter you found in the previous exercise. (*)

```{r}
surveyNames[ firstChar == mostPopularFirstChar ]
```

## Extra exercises {#extra}

1. Find the names that end with the most popular last character of all names.

```{r}
lastChar <- substr(surveyNames,nchar(surveyNames),nchar(surveyNames))
lastCharCounts <- table(lastChar)
( mostPopularLastChar <-  names(lastCharCounts[lastCharCounts==max(lastCharCounts)]) )
surveyNames[ lastChar == mostPopularLastChar ]
```

2. In the vector of e-mail addresses . 

Swap students between the two domains `lumc.nl` and `vumc.nl`. 

```{r}
swappedEmails <- emails
lumc <- grepl("lumc.nl",swappedEmails) 
vumc <- grepl("vumc.nl",swappedEmails)
swappedEmails[lumc] <- sub("lumc.nl","vumc.nl", swappedEmails[lumc])
swappedEmails[vumc] <- sub("vumc.nl","lumc.nl", swappedEmails[vumc])

#
# A more concise solution with *ifelse* (see ?ifelse)
#
swappedEmails2  <- ifelse(grepl("lumc.nl",emails), 
                          sub("lumc.nl","vumc.nl",emails), 
                          sub("vumc.nl","lumc.nl",emails)) 

identical(swappedEmails, swappedEmails2)
```

3. Some names are duplicated in the survey data set. Create a unique vector of e-mail addresses for these duplicated names by number suffixes, e.g. Barbra1, Barbra2 etc. 

```{r}
# First create table of non-unique names with counts as the number of occurences. 
tblOfNames<- table(surveyNames)
# Find the non-unique names
nonUniqueNames <- tblOfNames[ tblOfNames > 1 ]
# Make a copy of surveyNames called userNames and update the duplicated names. 
userNames <- surveyNames
# For each non-unique name find the names in userNames and replace them with 
# with <name>1 .. <name>n, where n is the number of occurences of
# the name in surveyNames.
for (u in names(nonUniqueNames)) {
  userNames[userNames==u] <- paste(u,1:nonUniqueNames[u],sep="")
}
# Create email addreses
emails2 <- paste(userNames, domains, sep="@")
```

4. Find top (most frequent) name initial for males and females.

```{r}
# Create a separate vector of male and female names.
femaleNames <- as.vector(subset(survey,gender=="female")$name)
maleNames   <- as.vector(subset(survey,gender=="male")$name)
# females initials
femalesLastChar <- substr(femaleNames,1,1)
femalesLastCharCounts <- table(femalesLastChar)
( femalesMostPopularLastChar <-  names(femalesLastCharCounts[femalesLastCharCounts==max(femalesLastCharCounts)]) )
femaleNames[ femalesLastChar == femalesMostPopularLastChar ]
# males initials
malesLastChar <- substr(maleNames,1,1)
malesLastCharCounts <- table(malesLastChar)
( malesMostPopularLastChar <-  names(malesLastCharCounts[malesLastCharCounts==max(malesLastCharCounts)]) )
maleNames[ malesLastChar == malesMostPopularLastChar ]
```


5. Order survey table by `name` character length (?order).

```{r}
# Ordered from short to long names. See ?order for more options.
surveyOrderedByNameLength <- survey[order(nchar(surveyNames)),] 
```


