---
title: "Character string processing & Pattern matching"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
#pulse <- read.table("https://barmsijs.lumc.nl/R_course/data/pulse.txt", header = T) 
#survey <- read.delim("https://barmsijs.lumc.nl/R_course/data/survey.txt") 
pulse <- read.table("../data/pulse.txt")
survey <- read.delim("../data/survey.txt")
source( "../R_course_setup.R" )
```

# Character string processing & Pattern matching

## Learning goals

- Character string manipulation : string concatenation,splitting, etc. 
- Pattern matching and replacement  

## Quotes and escape characters (characters with special meaning)

```
\n  newline                   \v	vertical tab
\r	carriage return           \\	backslash \
\t	tab                       \'	ASCII apostrophe '
\b	backspace                 \"	ASCII quotation mark "
\a	alert (bell)              \`	ASCII grave accent (backtick) `
\f	form feed                 \\.   dot (escaped twice ; special meaning in RE)
```

## Character string

Character string or simply `character`, as it is called in R, is a sequence of characters and a character vector is vector of character strings. 


```{r}
( cs <- "This is a character string" ) 
( cv <- c(cs, "and this is another !") )
```

Single and double quotes can be used interchangeably, however, double quotes are preferred.

```{r}
"string" 
'string' 
"'strings'"
'"string"'   
```

Formatting the output with `print` and `cat` functions. The function `print` does more formatting than `cat`. On the other hand `cat` interprets escape characters such as whitespaces [\\t\\n\\r,...], note that the behavior will differ depending on the platform. The function `cat` is useful and often used to print progress and/or debugging information in functions. 

```{r}
print(1:50)
cat(1:50)
print("separate\nlines")
cat("separate\nlines") 
cat("column1\tcolumn2") 

```

Some useful function:

```{r}
letters
LETTERS
month.name
month.abb
```

## paste(..., sep = " ", collapse = NULL) : 

Concatenates one or more vectors into a character vector.  

```{r}
paste("Bonnie","@", "lumc.nl")
paste("Travis","@", "lumc.nl", sep="")
paste("Travis", "lumc.nl", sep="@")
paste("Travis", 1,"@", "lumc.nl", sep="") # convert numeric to character 
paste("Travis","@","lumc.nl",";","Bonnie","@","lumc.nl", sep="")
paste(c("Bonnie","Travis"),"@", "lumc.nl", sep="") # recycling 
```

Recycling occurs in expressions involving multiple vectors of different sizes. The rule is that smaller vectors are recycled, partially if necessary, as often as possible to match the size of the largest vector.


```{r}

paste(c("Bonnie","Travis"),"@", "lumc.nl", sep="") 
paste(c("Bonnie","Travis"),"@", "lumc.nl", sep="", collapse = ";") 
paste(c("Bonnie","Travis"), "lumc.nl", sep="@", collapse = ";")
```

```{r}

```

## Pulse data set

```{r, echo=FALSE}
set.seed(12345)
```

```{r}
head(pulse) 
allNames <- as.vector(pulse$name)
head(allNames)
length(allNames)
( pulseNames <- sample(allNames, size = 5) ) # select randomly 5 names from allNames
domains <- c("lumc.nl", "leidenuniv.nl", "vumc.nl")
emails  <- paste(pulseNames,domains, sep="@") 
emails
```

## tolower, toupper, nchar

```{r}
toupper(emails) # convert to uppercase
tolower(emails) # convert to lowercase
nchar(emails)   # nr. of characters in string
```


```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "characterVectors", "character_strings.Rmd" )
```

## Split character string : strsplit(x, split, ...) : 

The `strsplit` function splits each element of the character vector `x` by substring `split` into substrings and returns a list of character vectors as the result.


```{r}
strsplit("Bonnie@lumc.nl", "@")
unlist( strsplit("Bonnie@lumc.nl", "@") )
strsplit(c("Bonnie@lumc.nl","Melanie@lumc.nl"), "@")
strsplit("Bonnie@lumc.nl", "n")
strsplit("Bonnie@lumc.nl", "") # useful
userDomains <- strsplit(emails,"@")
userDomains
```


## rbind(...), cbind(...) : combine by rows or columns

```{r,comment=NA}

letters[1:3]
LETTERS[1:3]
rbind(letters[1:3], LETTERS[1:3])
cbind(letters[1:3], LETTERS[1:3])
```

## strsplit(x, split, ...)


```{r}
userDomains <- strsplit(emails,"@")
userDomains
rbind(userDomains[[1]],userDomains[[2]],userDomains[[3]],userDomains[[4]],userDomains[[5]])
```


## do.call(what, args, ...) : execute function on list of arguments

```{r, comment=NA}
do.call(rbind, userDomains) # <=>  rbind(userDomains[[1]],userDomains[[2]],...)
```


```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "strsplit", "character_strings.Rmd" )
```

## grep(pattern, x, ignore.case = FALSE, value = FALSE, ... )

Search for matches of `pattern` in strings of character vector `x`.

```{r, comment=NA}
emails
grep("Dona", emails)
grep("Dona", emails, value = TRUE) 
grep("dona", emails, value = TRUE) # case-sensitive
grep("dona", emails, value = TRUE, ignore.case = TRUE)
grep("dona", emails, value = TRUE, ignore.case = TRUE, invert = TRUE)

```


## grepl(pattern, x, ignore.case = FALSE, ...) 

Same functionality as `grep` except it returns a logical vector of matches found. 

```{r, comment=NA}
emails
foundSubject <- grepl("Dona", emails)
foundSubject
emails[foundSubject]       # value
emails[ ! foundSubject ]   # invert
```


## substr(x, start, stop) : extract/replace substrings


```{r, comment=NA}
substr(x = "abc",start = 1,stop = 1)
substr("abc",1,nchar("abc"))
(abcs <- rep("abc",3))
substr(abcs,1,1:nchar("abc"))
(e <- head(emails))
substr(e,1,1) 
tolower(substr(e,1,1))
substr(e,1,1) <- tolower(substr(e,1,1)) # replace first character with its lowercase
e
```

## gsub/sub(pattern, replacement, x, ignore.case = FALSE, ...)

```{r, comment=NA}
sub("@", "(at)", emails)
sub("\\.", "\n", "git.lumc.nl")   # first occurence only 
gsub("\\.", "\n", "git.lumc.nl") # global: apply to all occurrences 
cat( gsub("\\.", "\n", "git.lumc.nl") )
```

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "grep", "character_strings.Rmd" )
```


## Extra exercises 

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "extra", "character_strings.Rmd" )
```


