---
title: "Basic data types (cont.)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup,include=FALSE,eval=TRUE}
source( "../R_course_setup.R" );
```

# Basic data types (cont.) 

## Missing values

Numeric, character, logical vectors, factors might contain elements marked as 'missing'.

`NA` is a constant which indicates a missing value.

`NA` values would appear in the course materials and tasks.

## Numeric vectors 

There are several ways to construct a vector of numbers.

### Single number (numeric vector with a single element)

```{r}
7
```

### Multiple numbers

```{r}
c( 5, -2.5, NA, 7+3, 1/3 )
```

### Sequence of numbers (one by one)

```{r}
3:17
```

### Sequence of numbers

Sequence of numbers with a defined step:
```{r}
seq( 5, 15, 2 )
```

### Combine (several) vectors of numbers

Multiple vectors can be combined together:
```{r}
v <- 1:9
w <- seq( 10, 90, 10 )
c( 0, v, w, 100 )
```

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "numeric", "01_basic_data_types_cont.Rmd" );
```

## Character vectors 

There are several ways to construct a vector of texts.

### Simple single text (character vector with a single element)

With double quotes:
```{r}
"Bioinformatics"
```

Or, with single quotes:
```{r}
'Biostatistics'
```

### Multiple texts

```{r}
c( "Bioinformatics", "Biostatistics" )
```

### Lines from a text file

Earlier we used `read.table` function to read a text file into a (two-dimensional) data frame. Here, use `readLines` function to read contents of the `pulse.txt` file into a (one-dimensional) character vector.
```{r include=FALSE}
txt <- readLines( "../data/pulse.txt" )
```
```{r eval=FALSE}
txt <- readLines( "data/pulse.txt" )
```

Each element of the vector `txt` is a single line from the file. The file is stored in `tsv` (tab-separated values format), so the `\t` symbol separates the values of columns. These are the first 3 lines (the first one contains the header of the table):
```{r}
head( txt, 3 )
```

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "character", "01_basic_data_types_cont.Rmd" );
```

## Logical vectors 

### Elementary logical values

```{r}
TRUE
FALSE
T
F
c( FALSE, F, TRUE, T )    # vector of logical values
```

(*) `NA` is a logical constant and it gets automatically converted to other types when necessary.

### Logical operators

#### Negation

Unary negation operator (denoted `!`):
```{r}
!TRUE
!FALSE
```

#### AND Operator `&`

Binary operator AND (denoted `&`) returns `TRUE` result when `all` its arguments are `TRUE`:
```{r}
TRUE & TRUE
```

Otherwise returns `FALSE`:
```{r}
FALSE & TRUE
FALSE & FALSE
```

#### OR operator `|`

Binary operator OR (denoted `|`) returns `TRUE` result when `any` (at least one) of its arguments is `TRUE`:
```{r}
TRUE | TRUE
FALSE | TRUE
```

Otherwise returns `FALSE`:
```{r}
FALSE | FALSE
```

### Relational operators

#### Equality operator `==`

```{r}
1 == 2
"Bioinformatics" == "Biostatistics"
FALSE == F
```

#### Inequality operators 

```{r}
1 != 2    # unequal  
"a" != "A"
FALSE != T
1 < 2     # less than
1 > 2     # greater than
2 <= 2    # less or equal 
2 >= 2    # greater or equal
```

### Comparison of two vectors

```{r}
v <- c( 0, 1, 2, 3, 4 )
w <- c( 4, 3, 2, 1, 0 )
v == w
v != w
v < w
v <= w
v >= w
```

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "logical", "01_basic_data_types_cont.Rmd" );
```

## Type conversions 

Sometimes a conversion to a vector of certain type might be needed.
The family of functions: `as.numeric`, `as.character`, `as.logical` take as an argument a vector of any type and return a vector of the type given in the function name.
When the conversion of an element is not possible, `NA` value is used.

```{r}
v <- 101:110
v
as.character( v )
```

```{r results="asis", echo=FALSE, comment=""}
catLinkTaskSection( "conversions", "01_basic_data_types_cont.Rmd" );
```

