---
title: "Exercises functions and apply"
date: "`r format(Sys.time(), '%d %B %Y')`"
---


```{r setup, include=FALSE}
source( "../R_course_setup.R" );
```

4. The data set USJudgeRatings has ratings for US superior court judges.

4.1 All variables except the first (CONT) are ratings between 1 and 10. Make a new data set by removing this first variable from the data set.

```{r}
USJ <- USJudgeRatings[,-1]
```

4.2 Make the correlation matrix of the different subratings. Hint: use cor().

```{r}
cor(USJ)
```

4.3 Two different ways of calculation of correlation are frequently used: the methods of Pearson and Spearman. Which correlation is calculated by default? Calculate the other type as well.

```{r,eval=FALSE}
?cor
```

```{r}
cor(USJ, method="spearman")
```

4.4 Calculate the average rating per judge. Hint: use rowMeans().

```{r}
rowMeans(USJ)
```

4.5 Calulate the number of unsatisfactory ratings (below 5.5) for each judge. Hint: first make a logical matrix of unsatisfactory TRUE/FALSE, and use rowSums().

```{r}
unsatisfactory <- USJ < 5.5
rowSums(unsatisfactory)
```

4.6 Calculate the median rating per judge. Hint: use median() and apply(). 

```{r}
apply(USJ, 1, median)
```

4.7 Calculate the standard deviation of the different subratings over the judges. 

```{r}
apply(USJ, 2, sd)
```


5. Let us reconsider the ChickWeight data, seen in exercise 2.

5.1 Calculate the mean weight for each diet type. Hint: use tapply().

```{r}
CW <- ChickWeight
tapply(CW$weight, CW$Diet, mean)
```


5.2 Calculate the number of chicks with weight over 300 grams for each diet type. Hint: use the function you wrote in exercise 2, together with tapply().

```{r echo=FALSE}
over300 <- function(x) {
  sum(x > 300)
}
```

```{r}
tapply(CW$weight, CW$Diet, over300)
```


6. The dataset Theoph has the results of an experiment on the pharmacokinetics of theophylline.

6.1 Make a line plot of time versus concentration for subject 1. Then add a line to the plot for time versus concentration of subject 2 (use the function lines()).

```{r}
plot(Theoph$Time[Theoph$Subject==1], Theoph$conc[Theoph$Subject==1], type="l")
lines(Theoph$Time[Theoph$Subject==2], Theoph$conc[Theoph$Subject==2])
```

6.2 Make a function that plots time versus concentration for subject i. This function takes the number i as input, and does not return anything.

```{r}
lines.i <- function(i) {
  lines(Theoph$Time[Theoph$Subject==i], Theoph$conc[Theoph$Subject==i])
}
```

6.3 Plot the data for subject 1 again, making sure the x and y-axis limits (xlim and ylim) are set to the full range of time and concentration for all subjects. Then use the function you created to add the lines for subject 2 up to 12 to the plot. Hint: use sapply().

```{r}
plot(Theoph$Time[Theoph$Subject==1], Theoph$conc[Theoph$Subject==1], type="l", xlim=range(Theoph$Time), ylim=range(Theoph$conc))
invisible(sapply(2:12, lines.i))
```

6.4 Make a color palette with 12 colors, e.g. using rainbow().

```{r}
palet <- rainbow(12)
```

6.5 Change your plotting function to let each line be drawn in its own color and plot again.

```{r}
lines.i <- function(i) {
  lines(Theoph$Time[Theoph$Subject==i], Theoph$conc[Theoph$Subject==i], col=palet[i])
}
plot(Theoph$Time[Theoph$Subject==1], Theoph$conc[Theoph$Subject==1], type="l", xlim=range(Theoph$Time), ylim=range(Theoph$conc), col=palet[1])
invisible(sapply(2:12, lines.i))
```

6.6 Optional: Make the same plot with a for-loop instead of an sapply.

```{r}
plot(Theoph$Time[Theoph$Subject==1], Theoph$conc[Theoph$Subject==1], type="l", xlim=range(Theoph$Time), ylim=range(Theoph$conc), col=palet[1])
for (i in 2:12) {
  lines.i(i)
}
```


7. The file tweedekamer2012.txt contains historical data on numbers of parliament seats of parties in the Dutch parliament from 1956 to date. Download the file to your computer.

7.1 Set the current working directory to the location where you downloaded the file tweedekamer2012.txt. Hint: use setwd().

7.2 Use read.table to read the file tweedekamer.txt. Make sure you set the argument quote="". Use tab as delimiter (check the help file of read.table() to see how to declare this.)

```{r}
tk <- read.table("tweedekamer2012.txt",quote="", sep="\t",header=T, row.names=1, check.names=F)
```

7.3 Check whether row and column names have been read correctly. Hint: use head. If necessary, adjust the arguments of read.table. Hint: look at check.names.

```{r}
head(tk)
```

7.4 How many parties and how many elections are in the data? Hint: use dim.

```{r}
dim(tk)
```

7.5 Try reading the file without using quote="". Find out what goes wrong by using read.table with the argument fill=TRUE.

```{r,eval=FALSE}
tk2 <- read.table("tweedekamer2012.txt", sep="\t", header=T, row.names=1, check.names=F, fill=TRUE)
```

7.6 Write the result to an R-readable .RData file using save.

```{r eval=FALSE}
save(tk, file="tweedekamer.RData")
```

7.7 Convert the data.frame read into a matrix.

```{r}
tk <- as.matrix(tk)
```

7.8 Create a numeric election year variable from the column names of your matrix. Hint: use as.numeric().

```{r}
year <- as.numeric(colnames(tk))
```

7.9 Calculate the number of parties represented in parliament after each election, and plot this number as a function of time as a line plot.

```{r}
# The number of parties represented in parliament after each election is the sum of entries in each column (year) that are larger than zero:
colSums(tk>0)
plot(year, colSums(tk>0), type="l")
```

7.10 Make the same plot as a barplot. Make the labels perpendicular to the axis for readability. Hint: see las in ?par.

```{r}
barplot(colSums(tk>0), las=2)
```

7.11 Find the names of all parties that have (so far) only been present in a single parliament.

```{r}
rownames(tk)[rowSums(tk>0)==1]
```

7.12 Find the name of the largest party in each parliament. Assign names to this object representing the year in which the parliament was elected.

```{r}
largest <- rownames(tk)[apply(tk, 2, which.max)]
names(largest) <- colnames(tk)
```

7.13 Plot the number of seats of the largest party as a function of time in a barplot. Color the bars by the name of the largest party, and make a proper legend.

```{r}
lseats <- apply(tk, 2, max)
barplot(lseats, las=2)
```



8.  Install the org.Hs.eg.db package from bioconductor.org. Use the BiocLite script. Hint: This is a large package: installing may take some time. If you start a second R session and do the installation there, you can continue working in R (on another exercise).

```{r,eval=FALSE,echo=TRUE}
source("http://bioconductor.org/biocLite.R")
biocLite("org.Hs.eg.db")
```

8.1 The org.Hs.eg.db package collects mappings from entrez gene to all kinds of genetic information. Use html help to view the mappings available.

8.2 Load the package using library.

```{r}
library(org.Hs.eg.db)
```

8.3 The mapping org.Hs.egCHR maps Entrez gene identifiers to chromosomes. What is the class of this object? In which package was this class defined? Find the vignette of this package.

```{r}
class(org.Hs.egCHR)
```

8.4 We can change the org.Hs.egCHR object to a familiar list object with as.list(). How many Entrez identifiers were mapped to chromosomes?

```{r}
chrom <- as.list(org.Hs.egCHR)
length(chrom)
```

8.5 On which chromosome is Entrez gene 57507 located?

```{r}
chrom["57507"]
```

8.6 Find the Gene symbol of Entrez gene 57507 using the org.Hs.egSYMBOL mapping.

```{r}
as.list(org.Hs.egSYMBOL)["57507"]
```

8.7 Check that org.Hs.egSYMBOL maps every Entrez identifier to at most one symbol. Hint use sapply, length, and all.

```{r}
# We first want to compute the length of each entry in the list as.list(org.Hs.egSYMBOL) - for this, we use sapply as in
sym.len <- sapply(as.list(org.Hs.egSYMBOL), length)
# The next step is to check which ones are equal to 1 - those correspond to genes (as defined by entrez identifiers) with a single symbol
sym.len1 <- sapply(as.list(org.Hs.egSYMBOL), length) == 1
# Finally, to check if all of them are equal to 1, we can use all() around this vector:
all(sapply(as.list(org.Hs.egSYMBOL), length) == 1)
```

8.8 Turn org.Hs.egSYMBOL into a character vector and remove all missing values. Hint: use unlist(), then na.omit().

```{r}
sym <- unlist(as.list(org.Hs.egSYMBOL))
sym <- na.omit(sym)
```

8.9 Find all gene symbols with "ESR"" in the name. Hint: use grep().

```{r}
ESRs <- grep("ESR", sym)
sym[ESRs]
```

8.10 On which chromosomes are these genes located?

```{r}
chrom[ESRs]
```



