---
title: "Regression (tasks)"
author:
- affiliation: "Bioinformatics Center of Expertise, Medical Statistics & Bioinformatics, LUMC"
  company: "LUMC"
  name: "Jelle Goeman"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

# Formulae and regression (tasks)

## Tasks

### Linear regression

We look at the `swiss` data. In this data set we explain fertility of Swiss regions using various indicators.

- Perform a linear regression with `Agriculture` as the explanatory variable. 

```{r}
res <- lm(Fertility ~ Agriculture, data=swiss)
summary(res)
```

- What is the confidence interval of the regression coefficient?

```{r}
confint(res, 'Agriculture')
```

- Perform another linear regession, but use a log-transformed `Agriculture`

```{r}
summary(lm(Fertility ~ log(Agriculture), data=swiss))
```

- Again, but use a `Agriculture` and `Agriculture` squared

```{r}
summary(lm(Fertility ~ Agriculture + I(Agriculture^2), data=swiss))
```

- Again, but use Agriculture, Catholic, and their interaction

```{r}
summary(lm(Fertility ~ Agriculture * Catholic, data=swiss))
```

- Make a variable `P` that has the p-value of the interaction in the last model. Hint: use coef of the summary of your object.

```{r}
fit <- lm(Fertility ~ Agriculture * Catholic, data=swiss)
coef(summary(fit))['Agriculture:Catholic', 'Pr(>|t|)']
```

- Is this model significantly better than the model with Agriculture alone? Hint: use anova.

```{r}
fit0 <- lm(Fertility ~ Agriculture, data=swiss)
anova(fit, fit0)
```

- Fit the model again in a data set that leaves out the canton `Courtelary`. Use the model to predict `Fertility` in `Courtelary`

```{r}
fit <- lm(Fertility ~ Agriculture * Catholic, data=swiss[-1,])
predict(fit, swiss[1,])
```

### Logistic regression

In the quine data (from MASS), we can try to predict learning speed.

- Make a logistic regression model to predict `Lrn` based on `Eth`, `Age` and `Sex`.

```{r}
library(MASS)
fit <- glm(Lrn ~ Eth + Age + Sex, data=quine, family=binomial)
```

- Is there a significant relationship between Age and Learning speed?
 
```{r}
fit0 <- glm(Lrn ~ Eth + Sex, data=quine, family=binomial)
anova(fit, fit0, test='LRT')
```



