---
title: "Good programming practice (GPP)"
author: ''
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
set.seed(1234)
```


```{r, eval=FALSE, include=FALSE, echo=FALSE, comment=NA}

## Notes

- everything is a function
  - what is the result of  typeof(if (FALSE) return("unreachable!"))
- example: m[i,j] <- if (…) … 

```


## Goal

- General coding conventions and style 
- User-defined Functions 

## Some general coding conventions 

To improve *readability* and *maintenance* easier:

  - File Names: end in .R
  - Identifiers: FunctionName, variableName, kConstantName
  - Line Length: maximum 70-80 characters
  - Indentation
  - Curly Braces: first on same line, last on own line
  - else: Surround else with braces
  - Assignment: use <-, not =
  - Semicolons: don't use them
  - General layout and ordering  (library, functions, main)
  - Use existing R functions if possible.
  - Write generic parts as functions
  - Inline documentation
  - Record sessionInfo() into project results. 

```{r, include=FALSE}
#- *Premature optimisation is the root of all evil  , Donald Knuth*

```

 
## R functions

```{r, echo=T, eval=TRUE}
intersect
class
typeof
```


## Running example : GC content

```{r, echo=FALSE}

strToChars <- function(x) {
  #
  # Description: converts a string to a character vector
  #
  unlist(strsplit(x,split=""))  
}

checkSequence <- function(sequence, sequenceType) {         
  #
  # Description: checks whether 'sequence' string
  # of type 'sequenceType' {DNA,RNA}
  #
  bases <- switch(sequenceType, 
                  DNA = "A|C|G|T", 
                  RNA = "A|C|G|U")
  lettersInAlphabet <- sum(grepl(bases, strToChars(toupper(sequence))))  
  
  return(lettersInAlphabet == length(strToChars(sequence)))
}

gcFrequency <- function(sequence, sequenceType=c("DNA", "RNA")) {
  # M. van Iterson 
  # June 3, 2015
  # input: 
  #        sequence DNA or RNA as a string
  #        sequenceType either 'DNA' or 'RNA'
  # output: 
  #        gc-frequency of the sequence
    
  sequenceType <- match.arg(sequenceType)       
  
  # extract nucleotides using CAPITALS
  SEQUENCE    <- toupper(sequence)
  nucleotides <- strToChars(SEQUENCE)

  # check if the sequence is a proper DNA or RNA sequence
  if(!checkSequence(sequence, sequenceType))
     stop(paste("Sequence:", sequence, " is not a ", sequenceType, " sequence !"))
      
  # calculated GC frequency
  seqLength <- length(nucleotides)
  gcCounts <- sum(grepl("C|G", nucleotides))  
  gcFreqs <- gcCounts/seqLength   
  
  return(gcFreqs)
}



```

- DNA/RNA are sequences of nucleobases with DNA={A,C,G,T} and RNA={A,C,G,U}
- GC frequency is the proportions of G's and C's in a sequence.

```{r}
( DNA1 <- paste(sample(c("A", "C", "G", "T"), 20, replace=TRUE), collapse="") )
( RNA1 <- paste(sample(c("A", "C", "G", "U"), 20, replace=TRUE), collapse="") )
```



```{r, echo=FALSE}
cat('DNA GC frequency:')
gcFrequency(DNA1, sequenceType = "DNA")
cat('RNA GC frequency:')
gcFrequency(RNA1, sequenceType = "RNA")
```


## GC content in a DNA sequence

```{r}
DNA1
( nucleotides <- unlist(strsplit(DNA1,"")) )
sum(grepl("[CG]", nucleotides))
sum(grepl("[CG]", nucleotides))/nchar(DNA1)

```


## gcFrequency function

```{r}
gcFrequency <- function(sequence) {
  seqLength   <- nchar(sequence)
  nucleotides <- unlist(strsplit(sequence,""))
  gcCounts    <- sum(grepl("[CG]", nucleotides))
  return(gcCounts/seqLength)
}

DNA1
gcFrequency(DNA1)

```


## Function arguments

```{r, error=TRUE}
( DNA2 <- paste(sample(c("a", "c", "g", "t"), 20, replace=TRUE), collapse="") ) 
gcFrequency(DNA2)
( DNA3 <- 12345 )
gcFrequency(DNA3) 
```

## stop(...)/warning(...)

When to issue:
 
 - **Warnings** : possibility of recovery and no potential harm to the end result
 
 - **Errors**   : no possibility for recovery and potential harm to the end result

```{r, error=TRUE, warning=TRUE}
stop("your error message !")
warning("your warning message !")
```


##  Function argument check

```{r, error=TRUE}

gcFrequency <- function(sequence) {
  sequence    <- as.character(sequence) 
  seqLength   <- nchar(sequence)
  nucleotides <- unlist(strsplit(sequence,""))
  # Check if the sequence is a proper DNA sequence
  if(sum(grepl("[ACGT]", nucleotides)) != nchar(sequence))
    stop(paste("The sequence:", sequence, " is not a DNA sequence !"))
  gcCounts <- sum(grepl("[CG]", nucleotides))
  gcCounts/seqLength
}


DNA1
gcFrequency(DNA1) 

DNA2
gcFrequency(DNA2)

DNA3
gcFrequency(DNA3)

```

## Function argument check (cont.)

```{r, error=TRUE}

gcFrequency <- function(sequence) {
  # Function 'toupper' coerces objects to character
  SEQUENCE    <- toupper(as.character(sequence))  
  seqLength   <- nchar(SEQUENCE)
  nucleotides <- unlist(strsplit(SEQUENCE,""))
  # Check if the sequence is a proper DNA sequence
  if(sum(grepl("[ACGT]", nucleotides)) != nchar(SEQUENCE))
    stop(paste("The sequence:", sequence, "is not a DNA sequence!"))
  gcCounts <- sum(grepl("[CG]", nucleotides))
  gcCounts/seqLength
}

DNA2 
gcFrequency(DNA2)

```


## Tasks

  1.  Update `gcFrequency` such that it can handle RNA sequences.
    - RNA sequences contain `Uracil` (U) instead of `Thymine` (T)
  2.  Modify the function, if necessary, so it can distinguish between DNA and RNA. Test it by an invalid input such as :  `uuuatataggguctggctat`. 


```{r, eval=FALSE, include=F}
#
#
#
gcFrequency <- function(sequence) {
  #
  # description: Takes a Character vector of length 1 representing 
  # DNA/RNA sequence and returns the GC proportion as the result. 
  #

  # Function 'toupper' coerces objects to character
  SEQUENCE    <- toupper(sequence)  
  seqLength   <- nchar(SEQUENCE)
  nucleotides <- unlist(strsplit(SEQUENCE,""))
  # Check if the sequence is a proper DNA sequence
  if(sum(grepl("[ACGTU]", nucleotides)) != nchar(SEQUENCE) |
     (sum(grepl("[U]", nucleotides))>0 & sum(grepl("[T]", nucleotides))>0)
     )
    stop(paste("The sequence:", sequence, "is not a DNA/RNA sequence!"))
  gcCounts <- sum(grepl("[CG]", nucleotides))
  gcCounts/seqLength 
}  
  
```





```{r, error=TRUE, eval=FALSE, include=FALSE}

## Function argument 'seuqenceType'


gcFrequency <- function(sequence, sequenceType=c("DNA", "RNA")) {
  
  # Check sequence type is  DNA or RNA
  sequenceType <- match.arg(sequenceType)
  bases <- c(DNA="[ACGT]")
  if(sequenceType != "DNA") 
    bases <- c(DNA="[ACGU]")
  
  SEQUENCE    <- toupper(sequence)
  seqLength   <- nchar(SEQUENCE)
  nucleotides <- unlist(strsplit(as.character(SEQUENCE),""))
  # Check if the sequence is a proper DNA or RNA sequence
  if(sum(grepl(bases, nucleotides)) != nchar(SEQUENCE))
    stop(paste("The sequence:", 
               sequence, "is probably not a", sequenceType, "sequence!"))
  # calculated GC frequency
  gcCounts <- sum(grepl("[CG]", nucleotides))
  return(gcCounts/seqLength) 
}

DNA2 <- paste(sample(c("A", "C", "G", "U"), 20, replace=TRUE), collapse="") 
gcFrequency(DNA2,sequenceType =  "DNA")
gcFrequency(DNA2,sequenceType =  "RNA")

## Homework

# Remark : A pre-condition is that `sequence` must be a non-empty string.
# 
# 1. Write a function `checkSequence(sequence, sequenceType)` that checks the input sequence given a sequence type and returns logical true in case of DNA or RNA and false otherwise.
# 
# 2. Update gcFrequency with the use of function `checkSequence`.
# 
# 3. Rewrite the the function checkSequence such that it returns a correct DNA or RNA sequence as a vector of nucleotides and a NULL in case of failure. Update gcFrequency with the new checkSequence function. 


## Notes on warnings or errors

# When to issue:
# 
# - **Warnings** : possibility of recovery and no potential harm to the end result
# 
# - **Errors**   : no possibility for recovery and potential harm to the end result
# 
# Resolve by:
# 
# - Make the right choice !
# 
# - try(...)/tryCatch(...) 

```

```{r, eval=FALSE, include=FALSE}

## Debugging : browser()

ln <- function(x) {
  browser()
  if (x<0)
      stop("x must be >= 0 !")
  else 
    log(x)
}
```

```{r , eval=FALSE, include=FALSE}
> browser()
Called from: top level 
Browse[1]> help
n          next
s          step into
f          finish
c or cont  continue
Q          quit
where      show stack
help       show help
<expr>     evaluate expression
```


```{r, eval=FALSE, include=FALSE}
## Debugging : debug()/undebug()

ln <- function(x) {
  if (x<0)
      stop("x must be >= 0 !")
  else 
    log(x)
}

debug(ln)
# run program containing ln(...)
undebug(ln)


```

