---
title: 'Assignment'
author:
- affiliation: "Bioinformatics Center of Expertise, Medical Statistics & Bioinformatics, LUMC"
  company: "LUMC"
  name: "Szymon M. Kielbasa, R. Monajemi"
date: "`r format(Sys.time(), '%d %B %Y')`"
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```


# Analyse `flights` data set {#final}

Use the following setting. You'll need to install `nycflights13` data package from `cran`. 

```{r warning=FALSE,message=FALSE,eval=TRUE, echo=TRUE}
library( dplyr )
library( tidyr )
library( ggplot2 )
# install.packages( "nycflights13" )
library( nycflights13 ) 
f <- flights
```

### Airplanes

The column `tailnum` (of `flights` table from `nycflights` library) can be used to identify an airplane. 
- How many different airplanes took of from New York in 2013?
```{r}
# Some tailnum's are missing, therefore we first filter out entries 
# with missing tailnum:
f <- f %>% filter( ! is.na(tailnum)  )
# (1) When distinct is used with no variables as argument then it 
# will take all variables in the dataframe, in this case the only 
# variable selected in the dataframe is tailnum: 
f %>% select( tailnum ) %>% distinct() %>% nrow()
# (2) Or use disctinct directly:
f %>% distinct( tailnum ) %>% nrow()
```

- Was there any airplane operated by more than one carrier?
```{r}
# (1) Find first the distinct combinations of airplanes and carriers 
# and then count the number of airplanes (tailnum) and finally select 
# counts >1. 
f %>% select( tailnum, carrier ) %>% distinct() %>% count( tailnum ) %>% 
  filter( n > 1 )
# (2) Or count the nr. of unique airplane and carrier combinations and 
# then count the number of unqiue airplanes. The first count will 
# introduce a new column 'n' in the dataframe therefore we need to 
# rename it to avoid conflict with the second count. 
f %>% count( tailnum, carrier ) %>% rename( airplaneCarrierPairs = n ) %>% 
  count( tailnum ) %>% filter( n > 1 )
```

- How many different airplanes were operated by each carrier? Which carrier operates the largest number of airplanes?
```{r}
# Similar as the previous but now we are interested in the frequency of carrier 
# occurencesin the unique combinations of airplane-carriers.  
f %>% count( carrier, tailnum ) %>% rename( airplaneCarrierPairs = n ) %>% 
  count( carrier ) %>% arrange( desc( n ) )
```

- Summarize the previous result in a barplot. Use `geom_bar( stat = "identity" )`. Reproduce: (*)
```{r eval=TRUE}
ggplot( 
  f %>% count( carrier, tailnum ) %>% rename( airplaneCarrierPairs = n ) %>% 
    count( carrier ) %>% arrange( desc( n ) ),
  aes( x = carrier, y = n ))    + 
  geom_bar( stat = "identity" ) +
  theme_bw() + ylab( "Number of operated airplanes (in NY)")
```

### Flight speeds?

The goal is to find mean flight speeds for each route.

- Add a `speed` column (calculated from `distance` and `air_time`)
```{r}
# Use the original flight data (including missing tailnum's)
f <- flights
f %>% mutate( speed = distance / air_time ) %>% select( origin, dest, speed, everything() ) 

```

- For each route calculate the mean speed (put it to column `meanSpeed`)

```{r}
# Some of the (origin,dest) pairs will result in speeds with NA as value. 
# This is due to missing air_time. You need to account for this with na.rm 
# argument to mean when calculating the means over the multiple speeds per 
# route with possible NA's 
f %>% mutate( speed = distance / air_time ) %>% group_by( origin, dest ) %>% 
  summarize( meanSpeed = mean( speed, na.rm = TRUE ) )
```

- Produce a histogram of the mean speeds. Question: what is the unit of speed? Reproduce: (*)
```{r eval=TRUE, warning=FALSE, message=FALSE}
ggplot( 
  f %>% mutate( speed = distance / air_time ) %>% group_by( origin, dest ) %>% 
    summarize( meanSpeed = mean( speed, na.rm = TRUE ) ),
  aes( x = meanSpeed ) ) +
  geom_histogram() +
  theme_bw()
```

- Maybe there is a relation of mean speed and distance? Produce a scatter plot of `meanSpeed` vs. `meanDistance`. Reproduce: (*)
```{r eval=TRUE, message=FALSE, warning=FALSE}
ggplot( 
  f %>% mutate( speed = distance / air_time ) %>% group_by( origin, dest ) %>% 
    summarize( meanSpeed = mean( speed, na.rm = TRUE ), meanDistance = mean( distance ) ),
  aes( x = meanDistance, y = meanSpeed )) +
  geom_point() +
  theme_bw()
```

<!-- - Try `ggplotly` version of the above plot. Add to `aes( ..., origin = origin, dest = dest )`. Hoover above points to see where the longest flights go (*) -->
```{r eval=FALSE, echo=FALSE}
p <- ggplot( 
  f %>% mutate( speed = distance / air_time ) %>% group_by( origin, dest ) %>% summarize( meanSpeed = mean( speed, na.rm = TRUE ), meanDistance = mean( distance ) ),
  aes( x = meanDistance, y = meanSpeed, origin = origin, dest = dest )
) +
  geom_point() +
  theme_bw()
# ggplotly( p )
```

### Names of airports of origin

- Study table `airports`, in particular, the columns `faa` and `name`. 

```{r}
# Inspect the dataframe airports:
glimpse(airports)
# We are interested in the airport codes (faa) and their names:  
( ap <- airports %>% select( faa, name ) ) 
```

- Identify all `faa` codes of airports of `origin`, note that the `origin` and `dest` variables in flights tables are FAA codes.  

```{r}
origin <- flights %>% select( origin ) %>% distinct()
origin
```

- Merge the above tables: to each airport of origin add its name, based on a match between `origin` and `faa` variables
```{r}
origin %>% rename( faa = origin ) %>% left_join( ap , by = "faa")
```
