---
title: 'S3 and S4 classes (tasks)'
date: "`r format(Sys.time(), '%d %B %Y')`"
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

# S3 and S4 classes (tasks) {#s3s4}

<!-- ## S3 and S4 classes {#s3s4} -->

Start a fresh workspace with no objects in your environment. You can do this either by running `rm(list=ls())` in your console or clicking the brush icon on the `Environment` pane (RStudio top right). Also restart the R session to remove any loaded packages (menu: Session -> Restart R).

Run the following line in your R console:

```{r, echo=TRUE, eval=FALSE}
source(url("https://barmsijs.lumc.nl/R_course/data/objects.R"))
```

```{r, echo=FALSE, message=FALSE}
# source(url("https://barmsijs.lumc.nl/R_course/data/objects.R"))
source("objects.R")
```

This will create a set of objects in your environment. Inspect the objects and answer the following questions: 

1. What is the class and the class category (S3/S4/base) of each object?

```{r}
library(pryr)
class(o1);otype(o1) 
class(o2);otype(o2) 
class(o3);otype(o3) 
class(o4);otype(o4) 
class(o5);otype(o5) 
```

2. List the slots of the objects if possible. 

```{r}
# o1 is an S4 class category so it has slots that can be retrieved by getSlots() 
getSlots(class(o1))
# o2 is an S3 so names() will retrieve the slots
names(o2)
# o3 is an S3 
names(o3)
# o4 is S3 
names(o4)
# o5 is a named vector, a base class, but has names as attribute so function 
# names() is still valid.  
names(o5)
```

3. Often class categories S3 and S4 have an slot called `call`. Find which objects have this attribute and report the value


```{r}
# The objects o2 and o3 have the slot 'call. 

# o2 : data used is sleep, see ?lm and ?sleep for more details.
o2$call

# o3 : is a density object on mtcars$hp , see ?density and ?mtcars
o3$call

```

4. Plot the objects. 

```{r}
plot(o1)
plot(o2)
plot(o3)
plot(o4)
plot(o5)
```



