---
title: 'Domestic flights from New York in 2013'
author:
- affiliation: "Biomedical Data Sciences, LUMC"
  company: "LUMC"
  name: "Ramin Monajemi"
- affiliation: "Biomedical Data Sciences, LUMC"
  company: "LUMC"
  name: "Szymon M. Kiełbasa"
date: "`r format(Sys.time(), '%d %B %Y')`"
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
library( tidyverse )
# install.packages( "nycflights13" )
library( nycflights13 )
# install.packages( "mapdata" )
library( mapdata )
source( "../R_course_setup.R" )
```

# Domestic flights from New York in 2013 {#final}

Use the following setting. You'll need to install `nycflights13` data package from `cran`. For the last extra task you need also `mapdata` package.

```{r, echo=TRUE}
library( tidyverse )
# install.packages( "nycflights13" )
library( nycflights13 )
# install.packages( "mapdata" )
library( mapdata )
```

## Flights data frames

`nycflights13` library provides several tables: `flights`, `airlines`, `airports`.
Have a quick look in the documentation (`?flights`, etc.) to get familiar with the variables of these tables. Use functions such as `str` and/or `glimpse` to display samples of the data.  

## What are the airports of origin?

Find all different airport identifiers (Federal Aviation Administration (FAA)) given in the `origin` variable of the `flights` table.

```{r}
flights %>%  distinct( origin )
```

```{r}
unique( flights$origin )
```

## Map airport information to flights.

Use the result from the previous task and join it with the airport information in `airports` table.

```{r}
flights %>% distinct( origin ) %>% rename( faa = origin ) %>% 
    left_join( airports, by = "faa" )
```

## Flights summary per destintion.

- Produce the following table:
    - columns:
        - `dest` - destination airport identification code
        - `numOfFlights` - total number of flights to the destination in 2013
        - `meanFlightsPerDay` - mean number of flights per day (round to 1 decimal).
- order with decreasing number of flights
- Select 5 most frequent destinations as a vector into `freqDest` object.

```{r}
flightSummary <- flights %>% 
  count( dest ) %>% 
  arrange( desc( n ) ) %>% 
  rename( numOfFlights = n ) %>% 
  mutate( meanFlightsPerDay = round( numOfFlights / 365, 1 ) )

flightSummary
```

```{r}
freqDest <- flightSummary %>% 
  head( 5 ) %>% 
  pull( dest )     # pull retrieves the given variable as a vector
freqDest
```

- What are the names and time zones of the top 5 destinations?

```{r}
airports %>% filter(faa %in% freqDest) %>% select(faa,name,tzone)
```


## Top 5 destinations flights frequency from NY

- Produce the following table:
    - variables: 
        - `dest` : name of a destination (see `freqDest` from previous exercise)
        - `EWR`, `JFK`, `LGA` (`faa` codes NY airports) with number of flights as values
    - Hint: i) create first a flight frequency table NY to top 5 destination (long table), ii) convert to wide table (see `spread` function tidyr package). 
    - set `NA` to zero (see ?spread )
    
    
```{r}
flights %>% 
  filter( dest %in% freqDest ) %>% 
  count( origin, dest ) %>% 
  spread( origin, n, fill = 0 )
```

## Airplanes and carriers

The column `tailnum` in `flights` table can be used to identify an airplane. How many different airplanes took off from New York in 2013?

```{r}
flights %>% 
  distinct( tailnum ) %>% 
  nrow()
```

Was there any airplane operated by more than one carrier? 

```{r}
flights %>% 
  distinct( tailnum, carrier ) %>% 
  count( tailnum ) %>% 
  filter( n > 1 )
```

Give an overview of the carriers and the number of airplanes they operated on in descending order. Use `airlines` table to add the carrier name.

```{r}

carriersAirplanes <- flights %>% 
  distinct( carrier, tailnum ) %>%
  count( carrier ) %>% 
  arrange( desc( n ) ) %>% 
  left_join( airlines, by = "carrier" )

carriersAirplanes
```

- Summarize the previous result in a barplot. Use `geom_bar( stat = "identity" )`.  (*)

```{r}
ggplot(carriersAirplanes, aes( x = carrier, y = n ))    + 
  geom_bar( stat = "identity" ) +
  theme_bw() + ylab( "Number of operated airplanes (in NY)")
```

## Flight speed

The goal is to find mean flight speed for each route.

- Create variable `flightSpeed` as a copy of `flights`:
    - add `distanceKM` column expressing `distance` in km (one mile is 1.609km)
    - add `airTimeHours` column expressing `air_time` in hours
    - add `speedKMH` column expressing speed in km/h
    
```{r}
flightSpeed <- flights %>% 
  mutate( distanceKm = distance * 1.609 ) %>% 
  mutate( airTimeHours = air_time / 60 ) %>% 
  mutate( speedKmh = distanceKm / airTimeHours ) %>% 
  select( origin, dest, speedKmh, everything() ) 
```

- How many flights have missing `speedKmh`? 

```{r}
flightSpeed %>% filter(is.na(speedKmh)) %>% nrow
```

- Group `flightSpeed` into routes (origin,destination). Add the columns `meanSpeedKmh` and `meanDistanceKm` for mean speed and distance respectively. Remember to `ungroup` the table. Put the result into variable `routeSpeeds`. Be careful to properly treat `NA` speeds.

```{r}
routeSpeeds <- flightSpeed %>% 
  group_by( origin, dest ) %>% 
  summarize( 
    meanSpeedKmh = mean( speedKmh, na.rm = TRUE ),
    meanDistanceKm = mean( distanceKm, na.rm = TRUE )
  ) %>% 
  ungroup()
routeSpeeds
```

- Produce a histogram of the mean speeds along different routes.  (*)

```{r}
ggplot( 
  routeSpeeds %>% filter(!is.nan( meanSpeedKmh) ), # summarise in 0 size group results into NaN
  aes( x = meanSpeedKmh ) ) +
  geom_histogram(bins=30) +
  theme_bw()
```

- Maybe there is a relation of mean speed and distance? Produce a scatter plot of `routeSpeedKmh` vs. `meanDistanceKm`. (*)

```{r}
ggplot( 
  routeSpeeds %>% filter(!is.nan( meanSpeedKmh) ),
  aes( x = meanDistanceKm, y = meanSpeedKmh, color = origin ) ) +
  geom_point( alpha = 0.5 ) +
  theme_bw()
```

## Fun: Where are the destinations

Understand and try the following code: (*)

```{r fig.width=12,fig.height=12,echo=TRUE }
world <- map_data( "world" )
usa <- map_data( "usa" )
dests <- airports %>% 
  semi_join( flights %>% rename( faa = dest ), by = "faa" )
ggplot() + 
  geom_polygon( data = world, aes( x = long, y = lat, group = group ) ) + 
  geom_polygon( data = usa, aes( x = long, y = lat, group = group ), color = "red" ) + 
  geom_point( data = dests, aes( x = lon, y = lat ), color = "orange" ) +
  coord_fixed(1.3)
```


